---
title: "Analysis"
author: "Florian Hartig"
date: "7/20/2017"
output: 
  html_document: 
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=F, warning=F}
load(file = "../data/cleaned.RData")

library(lme4)
library(lmerTest)
library(vegan)
library(betapart)
```

Notes: habitats are occasionally abbreviate with German words MR (Magerrasen) = dry grassland, FW (Fettwiese) = Hay Meadow. 

# Alpha diversity comparison 


## Taxonomic alpha 

First of all, we see a clear habitat pattern (Fig. 2 in main paper)

```{r}

plot(alpha ~ as.factor(Plot) , data = sites$alpha2, ylab = "Species richness", xlab = "", col = grey.colors(2)[as.numeric(Habitat)], las = 2, notch = T)

legend("topright", c("Dry grasslands", "Hay meadows"), pch = 15, col = grey.colors(2))

library(beeswarm)
beeswarm(alpha ~ as.factor(Plot) , data = sites$alpha2, add = T, cex = 0.5, pch = 16)

```


Regression to test for the dependence of alpha on all predictors

```{r}
alphaT = glmer(alpha ~ Habitat + scale(Distanz) + scale(`HA (m²)`) +  (1|Plot), data = sites$alpha2, family = poisson)
summary(alphaT)
```

Further plots to explore possible nonlinear confounding with other predictors


```{r}
par(mfrow = c(1,2))
plot(alpha ~ Distanz , data = sites$alpha2, col = as.numeric(Habitat)+1)
plot(alpha ~ log10(`HA (m²)`) , data = sites$alpha2, col = as.numeric(Habitat)+1)
```


Conclusion is that taxonomic diversity that the only significant factor associated with taxonomic alpha is habitat (Magerrasen has higher tax alpha). 


## Genetic alpha

Genetic alpha diversity (measured in SSWP/n-1) for the 6 species each in the dry and mesic grasslands against the frequency (FR) of subplots in which the species was found. The dashed line in each plot shows a linear regression estimate across all species in a given habitat, the solid lines show linear regression estimate per species. 

```{r}
par(mfrow = c(1,2))
plot(`SSWP/n-1` ~ FR , data = sites$genAlpha[sites$genAlpha$Habitat == "Magerrasen",], main = "Dry grasslands", pch = as.numeric(Art), xlab = "Frequency", ylab = "Genetic diversity [SSWP/n-1]", ylim = c(0,70) )

fit <- lm(`SSWP/n-1` ~ FR , data = sites$genAlpha[sites$genAlpha$Habitat == "Magerrasen",])
abline(fit, lty = 2)

species = unique(sites$genAlpha$Art[sites$genAlpha$Habitat == "Magerrasen"])
for(sp in species){
  xrange = range(sites$genAlpha$FR[sites$genAlpha$Art == sp])
  fit <- lm(`SSWP/n-1` ~ FR , data = sites$genAlpha[sites$genAlpha$Art == sp,])
  y1 = predict(fit, newdata = data.frame(FR = xrange[1]))
  y2 = predict(fit, newdata = data.frame(FR = xrange[2]))
  lines(xrange, c(y1, y2))
}

legend("topleft", legend = species, pch=as.numeric(species), cex = 0.8 )


plot(`SSWP/n-1` ~ FR , data = sites$genAlpha[sites$genAlpha$Habitat == "Fettwiese",], main = "Hay Meadows", pch = as.numeric(Art), xlab = "Frequency", ylab = "Genetic diversity [SSWP/n-1]", ylim = c(0,70))


fit <- lm(`SSWP/n-1` ~ FR , data = sites$genAlpha[sites$genAlpha$Habitat == "Fettwiese",])
abline(fit, lty = 2)

species = unique(sites$genAlpha$Art[sites$genAlpha$Habitat == "Fettwiese"])
for(sp in species){
  xrange = range(sites$genAlpha$FR[sites$genAlpha$Art == sp])
  fit <- lm(`SSWP/n-1` ~ FR , data = sites$genAlpha[sites$genAlpha$Art == sp,])
  y1 = predict(fit, newdata = data.frame(FR = xrange[1]))
  y2 = predict(fit, newdata = data.frame(FR = xrange[2]))
  lines(xrange, c(y1, y2))
}

legend("topleft", legend = species, pch=as.numeric(species), cex = 0.8 )
```

The comparison suggests that genetic diversity differs little within species, and apparent correlations between FR and genetic diversity occur because of a confounding of FR with species. To seperate these effects, we calculated linear mixed models, with population and species as random intercepts. 

```{r}
full <- lmer(`SSWP/n-1` ~  Habitat * scale(FR) + scale(`HA (m²)`) + scale(Distanz) + (1|Population) + (1|Art) ,  data = sites$genAlpha, REML = F)
summary(full)
```

Which showed that none of the factors are significant. Habitat is also not significant in a single regression, as long as species identity is included as a random intercept. 

```{r}
full <- lmer(`SSWP/n-1` ~ Habitat + (1|Population) + (1|Art),  data = sites$genAlpha, REML = F)
summary(full)
```

However, species is obviously a signficant predictor for both genetic diversity. The problem is that species is perfectly collinear with habitat, and closely colinear with FR, so that it's difficult to separate these effects. 

```{r}
fit <- lmer(`SSWP/n-1` ~  Art + (1|Population)  ,  data = sites$genAlpha)
anova(fit)
```

Both habitat and frequency would be siginficant when not controlling for species identity, but given the previous results, and the apparent collinearity between species identity and other factors, this seems statistically not sensible. 

```{r}
full <- lmer(`SSWP/n-1` ~  Habitat * scale(FR) + scale(`HA (m²)`) + scale(Distanz) + (1|Population)  ,  data = sites$genAlpha, REML = F)
summary(full)
```

In fact, it seems that both the habitat effects and the frequency effects in the latter regression are essentially driven by species identity, in that some species simply have a higher diversity, or at least there are not enough species to separate between an effect of species identity, and true FR / Habitat effects. 

To confirm that intutition, we additinally calculated a t-test between the average genetic diversity values per species for each habitat (n=6 for each group) 

```{r}
x = aggregate(`SSWP/n-1` ~ Art + Habitat, FUN = mean, data = sites$genAlpha)
t.test(`SSWP/n-1` ~ Habitat, data = x)
```

Which is n.s. 

The correlation between species 

```{r}
x1 = aggregate(`SSWP/n-1` ~ Art, FUN = mean, data = sites$genAlpha)
x2 = aggregate(FR ~ Art, FUN = mean, data = sites$genAlpha)
cor.test(x1$`SSWP/n-1`, x2$FR)
```



# Beta diversity comparison 


## Taxonomic beta


```{r, fig.width=10, fig.height=10}
heatmap(t(sites$community), Colv = NA,scale="column", col = terrain.colors(256))
```


```{r}


distSpecies = vegdist(sites$community, method="bray")

mod <- betadisper(distSpecies, sites$alpha$Habitat)
mod
anova(mod)
permutest(mod, pairwise = TRUE, permutations = 499)
plot(mod, ellipse = TRUE, hull = FALSE, conf = 0.90) # 90% data ellipse
boxplot(mod)
```


Mantel test for spatial structure

```{r}
mantel(as.matrix(distSpecies)[1:8,1:8], sites$geoDist[1:8,1:8], method="pearson", permutations=999)
mantel(as.matrix(distSpecies)[9:16,9:16], sites$geoDist[9:16,9:16], method="pearson", permutations=999)
```

Result: FW not significant with geographic distance, but MR is probably (stronger driven by environment)


## Genetic Beta

```{r, fig.width=10, fig.height=8}
par(mfrow = c(3,4))
for(i in 1:12) image(t(sites$genBeta[[i]]), col = terrain.colors(256))
```

Regression PhiPT gegen Habitat

```{r}
boxplot(PhiPT ~  Habitat ,  data = sites$genAlpha[1:12])
fit <- lm(PhiPT ~  Habitat ,  data = sites$genAlpha[c(1:6, 49:54),])
summary(fit)

t.test(sites$genAlpha$PhiPT[1:6], sites$genAlpha$PhiPT[49:54])

```

PhiPT slightly higher in MR

Mantel tests to test for spatial signal in genetic diversity

```{r}
# note: genetic data are ordered first the MR species, then the FW
distSpecies = vegdist(sites$community, method="bray")

outMR = rep(NA,6)
for (i in 7:12) outMR[i-6] = mantel(sites$genBeta[[i]], sites$geoDist[1:8,1:8])$signif
outMR

outFW = rep(NA,6)
for (i in 1:6) outFW[i] = mantel(sites$genBeta[[i]], sites$geoDist[9:16,9:16])$signif
outFW
```

Result: again, FW are all n.s., but so is MR (only one species significant), potentially slightly higher spatial effects in MR

# Taxonomic Gamma

```{r}
countsFW = apply(sites$community[1:8,],2, sum)
countsMR = apply(sites$community[9:16,],2, sum)
sum(countsFW > 0)
sum(countsMR > 0)
```

Result: FW has higher diversity
